full title

### 2020-10-12 paper/pldi21/detail.tex
Applied review

We formulate the concrete semantics of the core language as described in
Figure~\ref{fig:core-trans-rel}.  The transition relation between concrete
states is defined with the semantics of references and expressions using two
different forms \fbox{$\referrule{\st}{\refer}{\loc}$} and
\fbox{$\exprrule{\st}{\expr}{\val}$}, respectively.  The initial states are
$\istset = \{ (\ilab, \varnothing, \epsilon, \tladdr) \}$ where $\ilab$ denotes
the initial label, $\epsilon$ empty list, and $\tladdr$ the top-level
environment address.  The function $\labnext$ returns the next lable of the
given lable in the current program $\prog$.

@@

    \inferrule*[width=0.48\textwidth]
    {
      \abselem = (\absmem, \absctxt, \absaddr, \abscount)\\
      \prog(\lab) = \refer = \expr_f ( \expr_a )\\
      \referabssem{\refer}(\abselem) = L\\
      \fval{x}{\lab_b} \in \exprabssem{\expr_f}(\abselem)\\
      \exprabssem{\expr_a}(\abselem) = \absval_a\\
      \eabsaddr = \lab_b\\
      \eabsctxt = \absctxt[\eabsaddr \mapsto \absctxt(\eabsaddr) \cup \{
      (\absaddr, \lab, L) \} ]
    }
    {
      \viewtrans{\lab}{\lab_b}(\abselem) = (\absmem[(\eabsaddr, x) \mapsto
      \absval_a], \eabsctxt, \eabsaddr, \inc(\abscount, \eabsaddr))
    }

@@

    \inferrule*[width=0.48\textwidth]
    {
      \abselem = (\absmem, \absctxt, \absaddr, \abscount)\\
      A = \exprabssem{\expr_0}(\abselem) \cap \absaddrset\\
      S = \exprabssem{\expr_1}(\abselem) \cap \strset\\
    }
    {
      \referabssem{\expr_0 [ \expr_1 ]} = A \times S
    }
  \end{mathpar}

@@

In abstract semantics of the core language, we use the flow sensitivity with a
flow sensitive view abstraction $\fsviewmap: \labset \rightarrow \dom$ that
discriminates states using their labels: $\forall \lab \in \labset. \;
\fsviewmap(\lab) = \{ \st \in \stset \mid \st = (\lab, \_, \_, \_) \}$ Thus, the
sensitive abstract domain is defined as $\sabsdom = \labset \rightarrow
\absdom$.  We define an abstract state $\abselem \in \absdom$ as a tuple of an
abstract memory, an abstract context, an abstract address, and set of singleton
abstract addresses as follows:

@@

An abstract memory $\absmem \in \absmemset$ is a finite mapping from abstract
locations $\abslocset$ to abstract values $\absvalset$.  Abstract locations
$\abslocset$ are pairs of abstract addresses with variable names or string
values. Abstract addresses $\absaddrset$ are defined with the
\textit{allocation-site abstraction} that partitions concrete addresses
$\addrset$ based on their allocation sites $\labset$.  Abstract contexts
$\absctxtset$ are finite maps from abstract addresses to powersets of triples of
abstract addresses, views, and powerset of abstract locations.  For abstract
counting~\cite{abstract-gc-counting, revisit-recency} in static analysis, we
define abstract counters $\abscountset$ that maps from abstract addresses to
their abstract counts representing how many times each abstract address has been
allocated; $\abszero$ denotes that it have never been allocated, $\absone$ once,
and $\absmany$ more than or equals to twice.

@@

\subsection{Combined Analysis}

We define the symbolic states with the extended concrete values and the abstract
counters for addresses as follows:
\[
  \begin{array}{r@{~}c@{~}l}
    \symbstset &=& \labset \times \memset \times \ctxtset \times \eaddrset
    \times \abscountset\\
    \ctxtset &=& \addrset \finmap ((\addrset \times \labset \times \locset)
    \uplus \symbset)\\
    \valset &=& \pvalset \uplus \oaddrset \uplus \fvalset \uplus \symbset\\
    \abscountset &=& \addrset \rightarrow \{ \abszero, \absone, \absmany \}\\
  \end{array}
\]
We define the converter $\asconverter: \absdom \rightarrow \symbstset$ from
abstract states to corresponding symbolic states as follows:
\[
  \asconverter((\lab, (\absmem, \absctxt, \absaddr, \abscount))) = (\lab,
  \asconverter^\memset(\absmem), \asconverter^\ctxtset(\absctxt),
  \asconverter^\addrset(\absaddr), \abscount)\\
\]
The converter for memories $\asconverter^\memset$ is defined with the converter
for values $\asconverter^\valset$ where:
\[
  \asconverter^\valset(\absval) = \left\{
    \begin{array}{ll}
      \val & \text{if} \; \absval = \{ \val \}\\
      \absval & \text{otherwise}\\
    \end{array}
  \right.
\]
In a similar way, the converter for contexts $\asconverter^\ctxtset$ is defined
as follows:
\[
  \asconverter^\ctxtset(\absctxt)(\absaddr) = \left\{
    \begin{array}{ll}
      (\absaddr, \lab, \loc) & \text{if} \; \absctxt(\absaddr) = \{ (\absaddr,
      \lab, \{ \loc \}) \}\\
      \absctxt(\absaddr) & \text{otherwise}\\
    \end{array}
  \right.
\]
The opposite converter $\saconverter: \symbstset \rightarrow \absdom$ from
symbolic states to abstract states is defined as the inverse function of
$\asconverter$: $\saconverter = \asconverter^{-1}$.
The symbolic transition relation $\symbtrans$ is an extension of the original
concrete transition relation $\trans$.  When the exact values of symbolic values
are required, symbolic states have the symbolic transition relation with the
exception state $\excst$.  For example, we add the following rule for $\kwret$
statements:
\begin{mathpar}
  \inferrule
  {
    \prog(\lab) = \kwret \; \expr\\
    \exprrule{\st}{\expr}{\val}\\
    \ctxt(\addr) \in \symbset
  }
  {
    \st = (\lab, \mem, \ctxt, \addr)
    \trans \excst
  }
\end{mathpar}
We extend each rule of concrete semantics to support such behaviors of symbolic
values.

>>>>

We formulate the concrete semantics of the core language as described in
Figure~\ref{fig:core-trans-rel}.  The transition relation between concrete
states is defined with the semantics of references and expressions using two
different forms \fbox{$\referrule{\st}{\refer}{\loc}$} and
\fbox{$\exprrule{\st}{\expr}{\val}$}, respectively.  The initial states are
$\istset = \{ (\ilab, \varnothing, \epsilon, \tladdr) \}$ where $\ilab$ denotes
the initial label, $\epsilon$ empty map, and $\tladdr$ the top-level
environment address.  The function $\labnext$ returns the next lable of a
given lable in the current program $\prog$.

@@

    \inferrule*[width=0.48\textwidth]
    {
      \abselem = (\absmem, \absctxt, \absaddr, \abscount)\\
      \prog(\lab) = \refer = \expr_f ( \expr_a )\\
      \referabssem{\refer}(\abselem) = L\\
      \fval{x}{\lab_b} \in \exprabssem{\expr_f}(\abselem)\\
      \exprabssem{\expr_a}(\abselem) = \absval_a\\
      \eabsaddr = \lab_b\\
      \eabsctxt = \absctxt[\eabsaddr \mapsto \absctxt(\eabsaddr) \cup \{
      (\absaddr, \labnext(\lab), L) \} ]
    }
    {
      \viewtrans{\lab}{\lab_b}(\abselem) = (\absmem[(\eabsaddr, x) \mapsto
      \absval_a], \eabsctxt, \eabsaddr, \inc(\abscount, \eabsaddr))
    }

@@

    \inferrule*[width=0.48\textwidth]
    {
      \abselem = (\absmem, \absctxt, \absaddr, \abscount)\\
      A = \exprabssem{\expr_0}(\abselem) \cap \absaddrset\\
      S = \exprabssem{\expr_1}(\abselem) \cap \strset\\
    }
    {
      \referabssem{\expr_0 [ \expr_1 ]}(\abselem) = A \times S
    }
  \end{mathpar}

@@

In abstract semantics of the core language, we use the flow sensitivity with a
flow sensitive view abstraction $\fsviewmap: \labset \rightarrow \dom$ that
discriminates states using their labels: $\forall \lab \in \labset. \;
\fsviewmap(\lab) = \{ \st \in \stset \mid \st = (\lab, \_, \_, \_) \}$. Thus, the
sensitive abstract domain is defined as $\sabsdom = \labset \rightarrow
\absdom$.  We define an abstract state $\abselem \in \absdom$ as a tuple of an
abstract memory, an abstract context, an abstract address, and set of singleton
abstract addresses as follows:

@@

An abstract memory $\absmem \in \absmemset$ is a finite mapping from abstract
locations $\abslocset$ to abstract values $\absvalset$.  Abstract locations
$\abslocset$ are pairs of abstract addresses with variable names or string
values. Abstract addresses $\absaddrset$ are defined with the
\textit{allocation-site abstraction} that partitions concrete addresses
$\addrset$ based on their allocation sites $\labset$.  Abstract contexts
$\absctxtset$ are finite maps from abstract addresses to powersets of triples of
abstract addresses, views, and powerset of abstract locations.  For abstract
counting~\cite{abstract-gc-counting, revisit-recency} in static analysis, we
define abstract counters $\abscountset$ that maps from abstract addresses to
their abstract counts representing how many times each abstract address has been
allocated; $\abszero$ denotes that it has never been allocated, $\absone$ once,
and $\absmany$ more than or equals to twice.

@@

\subsection{Lazy Concrete Interpretation}

We define the symbolic states with the extended concrete values and the abstract
counters for addresses as follows:
\[
  \begin{array}{r@{~}c@{~}l}
    \symbstset &=& \labset \times \memset \times \ctxtset \times \eaddrset
    \times \abscountset\\
    \ctxtset &=& \eaddrset \finmap ((\eaddrset \times \labset \times \locset)
    \uplus \symbset)\\
    \valset &=& \pvalset \uplus \oaddrset \uplus \fvalset \uplus \symbset\\
    \abscountset &=& \oaddrset \rightarrow \{ \abszero, \absone, \absmany \}\\
  \end{array}
\]
We define the converter $\asconverter: (\viewset \times \absdom) \rightarrow
\symbstset$ from abstract states to corresponding symbolic states as follows:
\[
  \asconverter((\lab, (\absmem, \absctxt, \absaddr, \abscount))) = (\lab,
  \asconverter^\memset(\absmem), \asconverter^\ctxtset(\absctxt),
  \asconverter^\addrset(\absaddr), \abscount)\\
\]
The converter for memories $\asconverter^\memset$ is defined with the converter
for values $\asconverter^\valset$ where:
\[
  \asconverter^\valset(\absval) = \left\{
    \begin{array}{ll}
      \val & \text{if} \; \absval = \{ \val \}\\
      \absval & \text{otherwise}\\
    \end{array}
  \right.
\]
In a similar way, the converter for contexts $\asconverter^\ctxtset$ is defined
as follows:
\[
  \asconverter^\ctxtset(\absctxt)(\absaddr) = \left\{
    \begin{array}{ll}
      (\absaddr, \lab, \loc) & \text{if} \; \absctxt(\absaddr) = \{ (\absaddr,
      \lab, \{ \loc \}) \}\\
      \absctxt(\absaddr) & \text{otherwise}\\
    \end{array}
  \right.
\]
The opposite converter $\saconverter: \symbstset \rightarrow (\viewset \times
\absdom)$ from symbolic states to abstract states is defined as the inverse
function of $\asconverter$: $\saconverter = \asconverter^{-1}$.
The symbolic transition relation $\symbtrans$ is an extension of the original
concrete transition relation $\trans$.  When the exact values of symbolic values
are required, symbolic states have the symbolic transition relation with the
exception state $\excst$.  For example, we add the following rule for $\kwret$
statements:
\begin{mathpar}
  \inferrule
  {
    \prog(\lab) = \kwret \; \expr\\
    \exprrule{\st}{\expr}{\val}\\
    \ctxt(\addr) \in \symbset
  }
  {
    \st = (\lab, \mem, \ctxt, \addr)
    \trans \excst
  }
\end{mathpar}
We extend each rule of concrete semantics to support such behaviors of symbolic
values.

### 2020-10-12 paper/pldi21/formal.tex
Abstract semantics of core language
Fixed minor typo in formalization
Applied review

\subsection{Abstract Interpretation}
The abstract interpretation~\cite{abs-interp-1977, abs-interp-1992}
over-approximate the transfer $\transfer$ to the \textit{abstract transfer
function} $\abstransfer: \absdom \rightarrow \absdom$ to get the
\textit{abstract semantics} $\abssem{\prog}$ in finite iterations as follows:
\[
    \abssem{\prog} = \underset{n \rightarrow
    \infty}{\lim}{(\abstransfer)^n(\iabselem)}\\
\]
We define a \textit{state abstraction} $\dom \galois{\alpha}{\gamma} \absdom$ as
a Galois connection between the concrete domain $\dom$ and an abstract domain
$\absdom$ with a \textit{concretization function} $\gamma$ and a
\textit{abstraction function} $\alpha$.  The initial abstract state $\iabselem
\in \absdom$ represents an abstraction of the initial state set; $\ielem
\subseteq \gamma(\iabselem)$.  The abstract transfer function $\abstransfer:
\absdom \rightarrow \absdom$ is defined as $\abstransfer(\abselem) = \abselem
\join \absstep(\abselem)$ with an \textit{abstract one-step execution}
$\absstep: \absdom \rightarrow \absdom$.  For the sound state abstraction, the
join operator and the abstract one-step execution should satisfy the following
conditions:
\begin{itemize}
  \item $\forall \abselem_0, \abselem_1 \in \absdom. \; \gamma(\abselem_0) \cup
    \gamma(\abselem_1) \subseteq \gamma(\abselem_0 \join \abselem_1)$
  \item $\forall \abselem \in \absdom. \; \absstep \circ \gamma(\abselem) \subseteq
    \gamma \circ \absstep(\abselem)$
\end{itemize}

@@

% One of the most widely-used sensitivity techniques is \textit{flow sensitivity}
% using a flow sensitive view abstraction $\fsviewmap: \labset \rightarrow \dom$.
% It discriminates states using their labels as follows:
% \[
%   \forall \lab \in \labset. \; \fsviewmap(\lab) = \{ \st \in \stset \mid \st =
%   (\lab, \_) \}
% \]

@@

We define a \textit{lazy concrete interpretation} by extending the transition
relation $\trans$ as a symbolic transition relation $\symbtrans$ on symbolic
states. First, we extends the concrete states $\stset$ to \textit{symbolic
states} with symbolic values $\symbset$ that denote abstract values cannot be
concretized to singleton values. A symbolic transition relation $\symbtrans
\subseteq \symbstset \times (\symbstset \uplus \{ \excst \})$ is exactly same
with the original transition relation $\trans$ except when the relation requires
the exact value of symbolic values.  In this case, the symbolic state has the
relation a special exception state $\excst$ to represent that it is impossible
to interpret in a concrete way.  Thus, a symbolic state \textit{always} has a
symbolic transition relation with a single symbolic state, and the lazy concrete
interpret is linear execution until reaching the exception state.

To freely convert a pair of view and an abstract state to the corresponding
symbolic state and vice versa, we define two domain converters $\symbstset
\galois{\saconverter}{\asconverter} (\viewset \times \absdom)$.  In the
converter $\asconverter$ from abstract states to symbolic states, if an abstract
value $\absval$ represents a singleton value, the function transforms the
abstract value to the corresponding concrete value.  Otherwise, it converts the
abstract value as a symbolic value in the result symbolic state.  Two converters
convert given elements wihtout loss of information:
\[
  \saconverter \circ \asconverter = \asconverter \circ \saconverter = \identity
\]
where $\identity$ denotes the identity function.

@@

We define the \textit{combined analysis} of the abstract interpretation and the
lazy concrete interpretation.  In the combined analysis, we use the powerset of
symbolic states as the domain $\combdom = \powerset{\symbstset}$, and extend the
view transition as follows:
\[
  \begin{array}{l}
    \combviewtrans{\view}{\view'}(\combelem) =\\
    \qquad \{
      \symbst' \mid \symbst \in S \wedge
      \symbst \symbtrans \symbst' \wedge
      \saconverter(\symbst') = (\view', \_)
    \}\\
    \qquad \cup \left\{
    \begin{array}{ll}
      \{ \asconverter \circ \viewtrans{\view}{\view'}(\bigjoin D) \}
      & \text{if} \; D \neq \varnothing\\
      \varnothing & \text{otherwise}\\
    \end{array}
    \right.
  \end{array}
\]
where $S = \combelem\mid_{\checker}$ and $D =
\dot{\saconverter}(\combelem\mid_{\neg\checker})$.  The dot notation $\dot{f}$
denotes the element-wise extended function of a function $f$.  We could
configure when the lazy concrete interpretation or the abstract interpretation
is performed based on the definition of the $\checker$ function and its negation
$\neg\checker$.  It should satisfy the following condition for the soundness and
the termination of the combined analysis:
\[
  \checker(\symbst) \Rightarrow \symbst \symbtrans^k \excst \wedge
  1 \leq k \leq N
\]
where $N$ is the pre-defined maximum length of the lazy concrete interpretation.
The combined analysis just performs one-step symbolic transition for each
symbolic state that passes the filter $\checker$. On the other hand, it converts
remaining symbolic states to the corresponding abstract states, merges them into
a single abstract state, and performs the abstract one-step execution for the
abstract state.

>>>>

\subsection{Abstract Interpretation}
The abstract interpretation~\cite{abs-interp-1977, abs-interp-1992}
over-approximates the transfer $\transfer$ to the \textit{abstract transfer
function} $\abstransfer: \absdom \rightarrow \absdom$ to get the
\textit{abstract semantics} $\abssem{\prog}$ in finite iterations as follows:
\[
    \abssem{\prog} = \underset{n \rightarrow
    \infty}{\lim}{(\abstransfer)^n(\iabselem)}\\
\]
We define a \textit{state abstraction} $\dom \galois{\alpha}{\gamma} \absdom$ as
a Galois connection between the concrete domain $\dom$ and an abstract domain
$\absdom$ with a \textit{concretization function} $\gamma$ and an
\textit{abstraction function} $\alpha$.  The initial abstract state $\iabselem
\in \absdom$ represents an abstraction of the initial state set; $\ielem
\subseteq \gamma(\iabselem)$.  The abstract transfer function $\abstransfer:
\absdom \rightarrow \absdom$ is defined as $\abstransfer(\abselem) = \abselem
\join \absstep(\abselem)$ with an \textit{abstract one-step execution}
$\absstep: \absdom \rightarrow \absdom$.  For the sound state abstraction, the
join operator and the abstract one-step execution should satisfy the following
conditions:
\begin{itemize}
  \item $\forall \abselem_0, \abselem_1 \in \absdom. \; \gamma(\abselem_0) \cup
    \gamma(\abselem_1) \subseteq \gamma(\abselem_0 \join \abselem_1)$
  \item $\forall \abselem \in \absdom. \; \absstep \circ \gamma(\abselem) \subseteq
    \gamma \circ \absstep(\abselem)$
\end{itemize}

@@

@@

We define a \textit{lazy concrete interpretation} by extending the transition
relation $\trans$ as a symbolic transition relation $\symbtrans$ on symbolic
states. First, we extends the concrete states $\stset$ to \textit{symbolic
states} with symbolic values $\symbset$ that denote abstract values that cannot be
concretized to singleton values. A symbolic transition relation $\symbtrans
\subseteq \symbstset \times (\symbstset \uplus \{ \excst \})$ is exactly the same
with the original transition relation $\trans$ except when the relation requires
the exact value of symbolic values.  In this case, the symbolic state has the
relation a special exception state $\excst$ to represent that it is impossible
to interpret in a concrete way.  Thus, a symbolic state \textit{always} has a
symbolic transition relation with a single symbolic state, and the lazy concrete
interpret is linear execution until reaching the exception state.

To freely convert a pair of view and an abstract state to its corresponding
symbolic state and vice versa, we define two domain converters $\symbstset
\galois{\saconverter}{\asconverter} (\viewset \times \absdom)$.  In the
converter $\asconverter$ from abstract states to symbolic states, if an abstract
value $\absval$ represents a singleton value, the function transforms the
abstract value to the corresponding concrete value.  Otherwise, it converts the
abstract value as a symbolic value in the result symbolic state.  Two converters
convert given elements wihtout loss of information:
\[
  \saconverter \circ \asconverter = \asconverter \circ \saconverter = \identity
\]
where $\identity$ denotes the identity function.

@@

We define the \textit{combined analysis} of the abstract interpretation and the
lazy concrete interpretation.  In the combined analysis, we use the powerset of
symbolic states as the domain $\combdom = \powerset{\symbstset}$, and extend the
view transition as follows:
\[
  \begin{array}{l}
    \combviewtrans{\view}{\view'}(\combelem) =\\
    \qquad \{
      \symbst' \mid \symbst \in S \wedge
      \symbst \symbtrans \symbst' \wedge
      \saconverter(\symbst') = (\view', \_)
    \}\\
    \qquad \cup \left\{
    \begin{array}{ll}
      \{ \asconverter \circ \viewtrans{\view}{\view'}(\bigjoin D) \}
      & \text{if} \; D \neq \varnothing\\
      \varnothing & \text{otherwise}\\
    \end{array}
    \right.
  \end{array}
\]
where $S = \combelem\mid_{\checker}$ and $D =
\dot{\saconverter}(\combelem\mid_{\neg\checker})$.  The dot notation $\dot{f}$
denotes the element-wise extended function of a function $f$.  We could
configure when the lazy concrete interpretation or the abstract interpretation
is performed based on the definition of the $\checker$ function and its negation
$\neg\checker$.  It should satisfy the following condition for the soundness and
the termination of the combined analysis:
\[
  \checker(\symbst) \Rightarrow \symbst \symbtrans^k \excst \wedge
  1 < k \leq N
\]
where $N$ is a pre-defined maximum length of the lazy concrete interpretation.
The combined analysis just performs one-step symbolic transition for each
symbolic state that passes the filter $\checker$. On the other hand, it converts
remaining symbolic states to the corresponding abstract states, merges them into
a single abstract state, and performs the abstract one-step execution for the
abstract state.
